\documentclass[english]{article}
\usepackage[T1]{fontenc}
\usepackage[latin9]{inputenc}
\usepackage{color}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{esint}
\usepackage{babel}
\usepackage[round]{natbib}
\usepackage{color,hyperref}
\definecolor{darkblue}{rgb}{0.0,0.0,0.3}
\hypersetup{colorlinks,breaklinks,
linkcolor=darkblue,urlcolor=darkblue,
anchorcolor=darkblue,citecolor=darkblue}
\usepackage[capitalise, noabbrev]{cleveref}

\begin{document}
	\title{Solving PDEs Associated with Economic Models}
	\author{\large{\textsc{Matthieu Gomez \thanks{I thank Valentin Haddad, Ben Moll, and Dejanir Silva for useful discussions.}}}}
	\date{\today}
	\maketitle
	This package \href{https://github.com/matthieugomez/EconPDEs.jl}{EconPDEs.jl} introduces a fast and robust way to solve systems of PDEs + algebraic equations (i.e. DAEs) associated with economic models. This note details the underlying algorithm. 


\section{Write Finite Difference Scheme}
	The system of PDEs is written on a state space grid and derivatives are substituted by finite difference approximations. Importantly, first order derivatives are upwinded. This allows to naturally handle boudnary counditions at the frontiers of the state space. This also tends to make the scheme monotonous (for instance see \citet{achdou2014heterogeneous}). 
\section{Solve Finite Difference Scheme}
				
Denote $Y$  the solution and denote $F(Y)$ the finite difference scheme corresponding to a model. The goal is to find $Y$ such that $F(Y) = 0$. The package includes a solver especially written for these finite different schemes

\paragraph{Intuition.}
The existing literature in economics solves for $Y$ using using one of the two methods:

	\begin{enumerate}
		\item Non linear solver. The method solves for the non linear system $F (Y ) = 0$. A Newton-Raphson update takes the form
		\begin{align}\label{1}
			0 &= F(y_{t}) + J_{F}(y_t) (y_{t+1} - y_t)
		\end{align}
		The issue with this method is that it requires the initial guess to be sufficiently close to the solution.\footnote{This method is usded, for instance, by \citet{garleanu2015young}}
		\item ODE solver . The method solves for the ODE $F(Y) = \dot{Y}$. The solution of $F(Y)=0$ is obtained with $T\rightarrow +\infty$. \footnote{See, for instance, \citet{ditellabalance}, \citet{silva2015risk}.}
		With a simple explicit Euler method, updates take the form
		\begin{align}\label{2}
			0&= F(y_t) - \frac{1}{\Delta} (y_{t+1} -y_{t})
		\end{align}
		The issue with this method is that explicit time steps usually don't converge (i.e. time iterations are exploding).
	\end{enumerate}
	I propose to solve for $Y$ using a fully implicit Euler method.  Updates now take the form 
	\begin{align*}
		\forall t \leq T \hspace{1cm} 0&= F(y_{t+1}) - \frac{1}{\Delta}(y_{t+1} -y_{t})
	\end{align*}
	Each time step now requires to solve a non linear equation. I solve this non linear equation using a Newton-Raphson method. These inner iterations therefore take the form
	\begin{align}\label{inner}
		\forall i \leq I \hspace{1cm}	0 &= F(y_{t}^i) - \frac{1}{\Delta}(y_{t}^{i} -y_{t}) + (J_{F}(y_t^i) -  \frac{1}{\Delta})(y^{i+1}_{t} - y_t^i)
	\end{align}
	We know that the Newton-Raphson method converges if the initial guess is close enough to the solution. Since $y_{t}$ converges towards $y_{t+1}$ as $\Delta$ tends to zero, one can always choose $\Delta$ low enough so that the inner steps converge.Therefore, I adjust $\Delta$ as follows. If the inner iterations do not converge, I decrease $\Delta$. When the inner iteration converges, I increase $\Delta$. \par
	The update \cref{inner} can be see as weighted average of the Newton-Raphson step \cref{1} and of the explicit Euler step \cref{2}.  After a few sucessful implicit time steps, $\Delta$ is large and therefore the algorithm becomes like Newton-Rapshon. In particular, the convergence is quadratic around the solution. \par
	\paragraph{Pseudo Transient Method} 	This method is most similar to a method used in the fluid dynamics literature. In this context, it is called the Pseudo-Transient Continuation method, and is denoted $\Psi tc$. Formal conditions for the convergence of the algorithm are given in  \citet{kelley1998convergence}.\par
	The algorithm with $I=1$ and $\Delta$ constant corresponds to \citet{achdou2014heterogeneous}. They prove the convergence of this algorithm for models in partial equilibrium. Allowing $I > 1$ and adjusting $\Delta$  are important to ensure convergence in general equilibrium, which are non linear.

	\bibliography{bib}
	\bibliographystyle{aer}
\end{document}
